\section{Introduction}

The McDonald-Kreitman (MK) test \citep{mcdonald1991}
is among the most widely used methods for detecting natural selection
at the molecular level.
By comparing the ratio of nonsynonymous to synonymous changes
within species (polymorphism) to that between species (divergence),
the MK test provides a direct test of the neutral theory prediction
that these ratios should be equal. Departures from this expectation,
often quantified by the neutrality index \citep[NI;][]{rand1996}
or the fraction of adaptive substitutions $\alpha$ \citep{smith2002},
have revealed pervasive positive selection across a wide range of taxa
\citep[e.g.,][]{lozada-chavez2025, brown2024, liti2009, lam2010,
downing2009, hughes2008, li2008}.
% I put some examples here. Probably far more than we need, but rather have
% some we can pick and chose as needed.

A key challenge for the standard MK test is that
slightly deleterious nonsynonymous mutations,
which contribute to polymorphism but rarely fix,
inflate the number of nonsynonymous polymorphisms relative to
nonsynonymous substitutions and bias $\alpha$ downward.
Early approaches addressed this by excluding low-frequency
polymorphisms, either by distinguishing tip from interior
mutations on the gene tree \citep{templeton1996}
or by applying a derived allele frequency cutoff \citep{fay2001}.
The asymptotic MK test further addresses this limitation
by estimating $\alpha$ as a function of derived allele frequency
and extrapolating to fixation,
effectively removing the contribution of weakly deleterious variants
\citep{messer2013, haller2017}.
Rather than discarding low-frequency polymorphisms entirely,
the imputed MK test uses the ratio of low- to high-frequency
synonymous polymorphisms to estimate the expected number
of weakly deleterious nonsynonymous polymorphisms
and subtracts them from the total,
retaining more data while correcting for the bias
\citep{murga-moreno2022}.
To address lineage-specific selection, the polarized MK test
uses a second outgroup
to assign fixed differences to specific lineages,
allowing lineage-specific inference of selection \citep{begun2007}.

Despite the centrality of MK-based methods to molecular evolution,
existing software implementations remain fragmented.
% I have been formatting the names of software using monospaced font.
% I like it aesthetically, but disregard as needed.
\texttt{DnaSP} \citep{librado2009} provides the standard test
through a Windows-only graphical interface.
The \texttt{iMKT} R package and web server \citep{murga-moreno2019}
implements several extensions
but requires pre-computed summary statistics
and is tightly coupled to specific databases.
The \texttt{asymptoticMK} web tool \citep{haller2017}
implements the asymptotic test
but does not accept raw sequence data.
The \texttt{degenotate} software, part of the \texttt{snpArcher}
pipeline \cite{mirchandani2023}, can calculate both standard and 
imputed versions of the test; however it does so only using
genetic variants in VCF format and is thus dependent on
(and susceptible to) the liftover of outgroup sequences.
No existing tool provides a unified command-line interface
that takes aligned coding sequences as input
and supports the standard, polarized, and asymptotic MK tests
together with batch processing across thousands of genes
(\cref{tab:comparison}).

Here we present \mkado, a Python package that fills this gap.
\mkado takes codon-aligned FASTA files as input
and implements the standard, polarized, asymptotic,
and imputed MK tests,
as well as related statistics,
with a consistent command-line interface.
It supports parallel batch processing of genome-scale datasets,
automatic multiple testing correction,
and built-in visualization
including volcano plots and asymptotic $\alpha$ curves.
\mkado is installable via \texttt{pip}
and is designed to integrate naturally
with modern Python-based bioinformatics workflows.

\begin{table}[!t]
  \centering
  \caption{Feature comparison of MK test software. Checkmarks indicate
    built-in support; dashes indicate the feature is absent.}
  \label{tab:comparison}
  \vspace{6pt}
  \footnotesize
  \begin{tabular}{@{}lccccc@{}}
    \toprule
    Feature & \texttt{DnaSP} & \texttt{iMKT} & \begin{tabular}[c]{@{}c@{}}asymptotic\\[-3pt]MK\end{tabular} &
      \texttt{degenotate} & \mkado \\
    \midrule
    Sequence input        & \checkmark & web only   & ---          & ---        & \checkmark \\
    Standard MK           & \checkmark & \checkmark & ---          & \checkmark & \checkmark \\
    Polarized MK          & ---        & ---        & ---          & ---        & \checkmark \\
    Asymptotic MK         & ---        & \checkmark & \checkmark   & ---        & \checkmark \\
    Imputed MK            & ---        & \checkmark & ---          & \checkmark & \checkmark \\
    $\alpha_\text{TG}$    & ---        & ---        & ---          & ---        & \checkmark \\
    Frequency filtering   & ---        & \checkmark & \checkmark   & \checkmark & \checkmark \\
    CLI                   & ---        & ---        & ---          & \checkmark & \checkmark \\
    Python API            & ---        & ---        & ---          & ---        & \checkmark \\
    Batch processing      & ---        & ---        & ---          & ---        & \checkmark \\
    Parallel execution    & ---        & ---        & ---          & ---        & \checkmark \\
    Multiple testing corr.& ---        & ---        & ---          & ---        & \checkmark \\
    Visualization         & limited    & \checkmark & \checkmark   & ---        & \checkmark \\
    Open source           & ---        & \checkmark & \checkmark   & \checkmark & \checkmark \\
    Cross-platform        & ---        & \checkmark & \checkmark   & \checkmark & \checkmark \\
    \bottomrule
  \end{tabular}
\end{table}
% I added degenotate to the table above. To be sure, someone should do another pass
% at their docs to make sure we are not missing any of the options.
% I know Scott has used it in the pass, so we could check with him as wekk,.
