\section{Application}

To demonstrate \mkado at genome scales,
we assembled a dataset of 12,437 human protein-coding genes
with both polymorphism and divergence data.
Codon-aligned orthologous sequences
for human (\textit{Homo sapiens}),
chimpanzee (\textit{Pan troglodytes}),
and orangutan (\textit{Pongo abelii})
were obtained from \texttt{OrthoMaM} v12 \citep{orthomam2024}.
Human intraspecific polymorphism was drawn from 178 Yoruba (YRI)
individuals in the 1000 Genomes Project NYGC high-coverage dataset
\citep{1000genomes2015},
yielding 356 phased haplotypes per gene.
% Not for the paper, but it might be good to provide an example
% of how to do this in the docs, particularly if user are working
% on non-model systems without clean databases. I wrangled something 
% the barnacles, but we can clean it up and posted in the docs or
% even as a supplementary script in the main repo.
Variants were mapped onto the \texttt{OrthoMaM} alignment coordinates
to produce multi-species FASTA files
suitable for direct input to \mkado.

Using orangutan as the outgroup,
genome-wide pooled counts were
$D_n$\,=\,94,578, $D_s$\,=\,188,325, $P_n$\,=\,44,850,
and $P_s$\,=\,47,838.
The per-gene standard MK test identified numerous genes
with significant departures from neutrality,
including genes significant after Bonferroni correction
(\cref{fig:volcano}A). As expected,
the majority of testable genes showed NI~$>$~1,
reflecting the well-characterized excess
of slightly deleterious nonsynonymous polymorphism in humans
\citep{li2008}.
Using chimpanzee as the outgroup yielded fewer fixed differences
($D_n$\,=\,37,120, $D_s$\,=\,65,228), roughly 2.5-fold less than
orangutan (\cref{fig:chimp_mk}A),
and no individual genes survived Bonferroni correction,
consistent with the limited power of single-gene tests
at shorter divergence times.

The Tarone-Greenland weighted estimate \citep{stoletzki2011}
across all genes was strongly negative for both outgroups:
$\hat{\alpha}_\text{TG} = -0.853$
(95\% CI: $-0.888$ to $-0.812$)
for orangutan
and $\hat{\alpha}_\text{TG} = -0.651$
(95\% CI: $-0.682$ to $-0.607$)
for chimpanzee,
with corresponding $\text{NI}_\text{TG}$ values of 1.853 and 1.651.
These negative values confirm the pervasive excess
of weakly deleterious nonsynonymous polymorphism in humans.
Excluding singletons shifted both estimates upward
($\hat{\alpha}_\text{TG} = -0.794$ for orangutan;
$-0.606$ for chimpanzee),
as expected given the enrichment of slightly deleterious
variants at low frequency.
Applying a 15\% derived allele frequency cutoff
\citep{fay2001}
brought both estimates close to zero
($\hat{\alpha}_\text{TG} = -0.091$ for orangutan;
$-0.041$ for chimpanzee),
removing the bulk of low-frequency deleterious variants
but still failing to recover a signal of positive selection.

In contrast, the aggregated asymptotic MK test,
which models the frequency dependence of $\alpha$
and extrapolates to fixation \citep{messer2013},
successfully recovered a positive signal of adaptive evolution,
estimating $\hat{\alpha} = 0.163$
(95\% CI: 0.075 to 0.253; \cref{fig:volcano}B)
using orangutan as the outgroup,
indicating that approximately 16\% of amino acid substitutions
on the human lineage were driven by positive selection.
With chimpanzee as the outgroup the estimate widened to
$\hat{\alpha} = 0.075$ (95\% CI: $-1.22$ to $0.11$;
\cref{fig:chimp_mk}B),
illustrating the greater uncertainty
from the shallower divergence.
These results are consistent with previous analyses
of human coding sequences \citep{messer2013}.

The aggregated imputed MK test \citep{murga-moreno2022},
using a 15\% derived allele frequency cutoff
and orangutan as the outgroup,
estimated $\hat{\alpha} = -0.097$,
while the chimpanzee outgroup gave
$\hat{\alpha} = -0.068$.
Like the frequency-filtered Tarone-Greenland estimates,
the imputed correction substantially reduces the downward bias
from weakly deleterious polymorphism
but does not recover a positive signal of adaptation,
consistent with the known difficulty of detecting positive selection
in human protein-coding genes without modeling
the full frequency dependence of $\alpha$.

The entire analysis---standard MK tests,
Tarone-Greenland estimation, asymptotic estimation,
imputed estimation,
and volcano plot generation for all 12,437 genes---completed
in under one minute on a multi-core workstation
using \mkado's parallel batch mode.
